Reconstructing the neutron-star equation of state with gravitational-wave detectors from a realistic population of inspiralling binary neutron stars

### 2014-10-31 paper.tex
Made change to Averaged likelihood subsection in response to Will Farr's comment, but I still think the justification for the 0-noise is needed, since this paper is not only directed at GW parameter estimation people.

%%%%%%%% Old version %%%%%%%%%

%Given the GW model and EOS fit described in Sections~\ref{sec:waveform} and~\ref{sec:eos}, we will describe here a method for estimating the EOS parameters from a set of GW observations. We use Bayes' theorem to write the posterior density function (PDF) for the universal EOS parameters $\vec E = \{\log(p_1), \Gamma_1, \Gamma_2, \Gamma_3\}$ that are common to all NSs~\cite{Glendenning1996book} and the waveform parameters $\vec\theta = \{\les{d}{}, \alpha, \delta, \psi, \iota, t_c, \phi_c, m_1, m_2, \tilde\Lambda, \delta\tilde\Lambda\}$ that are unique to each of the $n$ BNS systems as
%\begin{equation}
%\begin{split}
%p(\vec E,\vec\theta_1,\dots,\vec\theta_n | D,\mathcal{H},\mathcal{I})
%&= p(\vec E,\vec\theta_1,\dots,\vec\theta_n | \mathcal{H},\mathcal{I}) \\
%& \times\frac{ p(D | \vec E,\vec\theta_1,\dots,\vec\theta_n,\mathcal{H},\mathcal{I}) }{ p(D | \mathcal{H},\mathcal{I}) }.
%\end{split}
%\end{equation}
%The data $D = \vec d_1,\dots, \vec d_n$ is composed of the data streams $\vec d_i(t)$ from the GW detector network for each of the $n$ inspiral events. $\mathcal{H}$ represents the \ben{EOS parameterization with unknown parameters $E$ and the waveform model, chosen here to be the 4-parameter piecewise polytrope and the TaylorF2 waveform model respectively.}{EOS and waveform models, and} $\mathcal{I}$ represents the background information for the \ben{EOS and waveform}{} parameters. The quantity $p(\vec E,\vec\theta_1,\dots,\vec\theta_n | \mathcal{H},\mathcal{I})$ is the prior probability density for the EOS and waveform parameters, $p(D | \vec E,\vec\theta_1,\dots,\vec\theta_n,\mathcal{H},\mathcal{I})$ is the likelihood, and the normalization constant $p(D | \mathcal{H},\mathcal{I})$, which we will not need to calculate here, is the evidence. The marginalized PDF for the EOS parameters is then
%\begin{equation}
%\label{eq:margEOS}
%p(\vec E | D,\mathcal{H},\mathcal{I}) = \int d\vec\theta_1 \dots d\vec\theta_n p(\vec E,\vec\theta_1,\dots,\vec\theta_n | D,\mathcal{H},\mathcal{I}).
%\end{equation}
%This $11n$-dimensional integral is not easily computed, so we will decompose it into blocks that can be individually computed. First, we decompose the waveform parameters into extrinsic/nuisance parameters $\vec\theta_{\rm ex} = \{\les{d}{}, \alpha, \delta, \psi, \iota, t_c, \phi_c, \delta\tilde\Lambda\}$\footnote{$\delta\tilde\Lambda$ is an intrinsic parameter that in principle provides additional EOS information. However, it is unmeasurable with aLIGO~\cite{WadeCreightonOchsner2014}, so we treat it as a nuisance parameter and group it with the extrinsic parameters when we marginalize over them.} that are unrelated to the EOS measurement and intrinsic parameters $\vec\theta_{\rm in} = \{m_1, m_2, \tilde\Lambda\}$ that are relevant for measuring the EOS such that $\vec\theta = \{\vec\theta_{\rm ex}, \vec\theta_{\rm in}\}$.

%%%%%%%%%%%%%%%%%%

@@

In order to determine the characteristic ability of GW detectors to measure EOS parameters, we will want to average our results for Eq.~\eqref{eq:margEOS2} over many realizations of the detector noise. In Ref.~\cite{NissankeHolzHughes2010} it was shown that averaging over noise realizations is equivalent to setting the noise $n^a(t)$ used to generate the data $d^a(t)$ equal to zero, but still using the characteristic noise PSD in the expression for the likelihood. Explicitly, one can define the averaged likelihood as the geometric mean of an ensemble of $M$ identical detectors measuring the same GW event
\begin{equation}
p_{\rm ave}(d^a | \vec\theta, \mathcal{H}, \mathcal{I}) = \left[\prod_{k=1}^M p(d_k^a | \vec\theta, \mathcal{H}, \mathcal{I})\right]^{1/M}.
\end{equation}
Taking $M$ to be large, it was shown that
\begin{equation}
\label{eq:averagedlike}
p_{\rm ave}(d^a | \vec\theta, \mathcal{H}, \mathcal{I}) \propto e^{-\frac{1}{2}(m^a(\vec\theta) - h^a(\hat\theta) | m^a(\vec\theta) - h^a(\hat\theta))}.
\end{equation}
This averaged likelihood will allow us to examine characteristic measurement uncertainties in the EOS parameters independent of the particular noise realization. It will also allow us in Section~VI to separate the effects of systematic errors due to uncertainties in the waveform model from effects due to individual noise realizations.

@@

%\ben{[[Why is the aVirgo PSD a few times higher between 500--1000Hz than the aLIGO one in the reference (Fig. 1)? I've seen noise curves where the high frequency is more comparable. Do you know of a reference that describes the different aVirgo configurations and noise curves?]]}{} \les{(I don't know of a reference for different aVirgo configurations. Here's a link to the PSDs in LAL: \url{https://www.lsc-group.phys.uwm.edu/daswg/projects/lal/nightly/docs/html/_l_a_l_sim_noise_p_s_d_8c.html}. Maybe this is helpful?)}{}

@@

%\red{[[Above $1.93M_\odot$, we do not expect the true $R(M)$ and $\lambda(M)$ curves to be contained in the credible region. This is because plotting the curve above $1.93M_\odot$ assumes that an object of that mass is a NS, whereas the credible regions allow for the possibility that the object is either a BH or a NS. If we assumed that a NS was a stable configuration above $1.93M_\odot$, this would be changing the prior.]]}

\begin{figure*}[!htb]
\begin{center}
\includegraphics[width=3.2in]{LALMCMCmpa1FitTaylorF2ZeroNoiseBNS20p.pdf}
\includegraphics[width=3.2in]{LALMCMCmpa1FitTaylorF2ZeroNoiseBNS20Radiuslambda.pdf}
%\includegraphics[width=2.5in]{LALMCMCmpa1FitTaylorF2ZeroNoiseBNS20RadiuslambdaHighMass.pdf}
\caption{Uncertainty in the recovered EOS, radius, and tidal deformability for the three-detector aLIGO--aVirgo network. Results are shown for the loudest 20 events with network SNR from 63.7 down to 13.6. The red, green, and blue regions represent the 68\% ($1\sigma$), 95\% ($2\sigma$), and 99.7\% ($3\sigma$) credible regions respectively. In the bottom left panel, $p_{\rm true}$ is the pressure of the ``true'' injected EOS, which in this case is the fit to the MPA1 EOS. In the right panels, the vertical line at $1.93M_\odot$ is the minimum mass, set by the prior, where some accepted EOS parameters do not produce a stable NS.}
\label{fig:mpa1fit}
\end{center}
\end{figure*}

@@

%[[Mode coupling. Tidally deformed NSs are unstable to the coupling of high order p-modes and g-modes~\cite{WeinbergArrasBurkart2013}. Accounting for three- and four-mode couplings, this instability cancels~\cite{VenumadhavZimmermanHirata2014}.]]

@@

BL sincerely thanks Walter Del Pozzo and Tyson Littenberg for helpful advice and corrections, as well as Will Farr, Tjonnie Li, and Chris Van Den Broeck. BL had several useful discussions at the University of Washington's Institute for Nuclear Theory Program 14-2a. The authors would also like to thank Evan Ochsner and Jolien Creighton for suggestions and help in generating the parameters of the BNS populations used here. This work was supported by NSF Grants PHY-1305682, PHY-0970074, and PHY-0955929. Simulations were run on the Nemo cluster at the University of Wisconsin--Milwaukee supported by the NSF Grant PHY-0923409, and the Orbital cluster at Princeton University.

>>>>

@@

%In order to determine the characteristic ability of GW detectors to measure EOS parameters, we will want to average our results for Eq.~\eqref{eq:margEOS2} over many realizations of the detector noise. In Ref.~\cite{NissankeHolzHughes2010} it was shown that averaging over noise realizations is equivalent to setting the noise $n^a(t)$ used to generate the data $d^a(t)$ equal to zero, but still using the characteristic noise PSD in the expression for the likelihood. Explicitly, one can define the averaged likelihood as the geometric mean of an ensemble of $M$ identical detectors measuring the same GW event
%\begin{equation}
%p_{\rm ave}(d^a | \vec\theta, \mathcal{H}, \mathcal{I}) = \left[\prod_{k=1}^M p(d_k^a | \vec\theta, \mathcal{H}, \mathcal{I})\right]^{1/M}.
%\end{equation}
%Taking $M$ to be large, it was shown that
%\begin{equation}
%\label{eq:averagedlike}
%p_{\rm ave}(d^a | \vec\theta, \mathcal{H}, \mathcal{I}) \propto e^{-\frac{1}{2}(m^a(\vec\theta) - h^a(\hat\theta) | m^a(\vec\theta) - h^a(\hat\theta))}.
%\end{equation}
%This averaged likelihood will allow us to examine characteristic measurement uncertainties in the EOS parameters independent of the particular noise realization. It will also allow us in Section~VI to separate the effects of systematic errors due to uncertainties in the waveform model from effects due to individual noise realizations.

In order to determine the characteristic ability of GW detectors to measure EOS parameters, we will want to average our results for Eq.~\eqref{eq:margEOS2} over many realizations of the detector noise. Ref.~\cite{NissankeHolzHughes2010} defined the averaged likelihood as the geometric mean of an ensemble of $M$ identical detectors measuring the same GW event
\begin{equation}
p_{\rm ave}(d^a | \vec\theta, \mathcal{H}, \mathcal{I}) = \left[\prod_{k=1}^M p(d_k^a | \vec\theta, \mathcal{H}, \mathcal{I})\right]^{1/M}.
\end{equation}
Taking $M$ to be large, they found that this is equivalent to setting the noise $n^a(t)$ used to generate the data $d^a(t)$ equal to zero, but still using the characteristic noise PSD in the expression for the likelihood:
\begin{equation}
\label{eq:averagedlike}
p_{\rm ave}(d^a | \vec\theta, \mathcal{H}, \mathcal{I}) \propto e^{-\frac{1}{2}(m^a(\vec\theta) - h^a(\hat\theta) | m^a(\vec\theta) - h^a(\hat\theta))}.
\end{equation}
This zero-noise likelihood will allow us to examine characteristic measurement uncertainties in the EOS parameters independent of the particular noise realization. It will also allow us in Section~VI to separate the effects of systematic errors due to uncertainties in the waveform model from effects due to individual noise realizations.

@@

@@

\begin{figure*}[!htb]
\begin{center}
\includegraphics[width=3.2in]{LALMCMCmpa1FitTaylorF2ZeroNoiseBNS20p.pdf}
\includegraphics[width=3.2in]{LALMCMCmpa1FitTaylorF2ZeroNoiseBNS20Radiuslambda.pdf}
\caption{Uncertainty in the recovered EOS, radius, and tidal deformability for the three-detector aLIGO--aVirgo network. Results are shown for the loudest 20 events with network SNR from 63.7 down to 13.6. The red, green, and blue regions represent the 68\% ($1\sigma$), 95\% ($2\sigma$), and 99.7\% ($3\sigma$) credible regions respectively. In the bottom left panel, $p_{\rm true}$ is the pressure of the ``true'' injected EOS, which in this case is the fit to the MPA1 EOS. In the right panels, the vertical line at $1.93M_\odot$ is the minimum mass, set by the prior, where some accepted EOS parameters do not produce a stable NS.}
\label{fig:mpa1fit}
\end{center}
\end{figure*}

@@

@@

We sincerely thank Walter Del Pozzo, Tyson Littenberg, and Jolien Creighton for advice and corrections, as well as Evan Ochsner for help in generating the parameters of the BNS populations used here. BL also benefited from discussions with Will Farr, Tjonnie Li, and Chris Van Den Broeck and from discussions at the University of Washington's Institute for Nuclear Theory Program 14-2a. This work was supported by NSF Grants PHY-1305682, PHY-0970074, PHY-0955929, and AST-1333142. Simulations were run on the Nemo cluster at the University of Wisconsin--Milwaukee, supported by the NSF Grant PHY-0923409, and the Orbital cluster at Princeton University.
